---
title: "Introduction to rosace"
author: "Jingyou Rao"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rosace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(tidy = FALSE, cache = FALSE, dev = "png",
                      message = FALSE, error = FALSE, warning = TRUE)
```	

## Installation

To install __rosace__ start [R](https://www.r-project.org) and first install devtools by typing
```{r eval=FALSE}
install.packages("devtools")
```

and install __rosace__ by typing
```{r eval=FALSE}
devtools::install_github("pimentellab/rosace")
```

If downloading from local type
```{r eval=FALSE}
devtools::install(".") # change to the directory of rosace
```

Next load __rosace__ with
```{r}
library("rosace")
```

## Example

#### Read Count Data

To explain how to use __rosace__ we provide an example based on the OCT1 drug cytotoxicity screen (https://www.biorxiv.org/content/10.1101/2023.06.06.543963v1.full). 

The screen has three replicates. The count matrix is stored in "data/oct1.rda".

```{r}
data("oct1")
key <- "1SM73"
type <- "growth"
```

#### Create Rosace Object

First, load the count file into the Assay object. 
Each replicate in the experiment will form an Assay object and will share the same "key" (1SM73).

```{r}
assay1 <- CreateAssayObject(counts = as.matrix(oct1_rep1[2:ncol(oct1_rep1)]),
                            var.names = oct1_rep1$hgvs,
                            key = key, rep = 1, type = type)
assay2 <- CreateAssayObject(counts = as.matrix(oct1_rep2[2:ncol(oct1_rep2)]),
                            var.names = oct1_rep2$hgvs,
                            key = key, rep = 2, type = type)
assay3 <- CreateAssayObject(counts = as.matrix(oct1_rep3[2:ncol(oct1_rep3)]),
                            var.names = oct1_rep3$hgvs,
                            key = key, rep = 3, type = type)
```

Next create Rosace object by adding three Assay objects together.

```{r}
rosace <- CreateRosaceObject(object = assay1)
rosace <- AddAssayData(object = rosace, assay = assay2)
rosace <- AddAssayData(object = rosace, assay = assay3)
GetAssayName(rosace)
```

#### Preprocessing

'CreateAssayObject' calls function 'FilterData' to filter the variants with more than 'na.rmax'% of NAs by default. But we might want to filter more variants later.

```{r}
rosace <- FilterData(rosace, key = key, na.rmax = 0.5)
```

Then we will impute the NA data either by K-Nearest Neighbor method or fill the NA with 0. 

```{r}
rosace <- ImputeData(rosace, key = key, impute.method = "knn", na.rmax = 0.5)
# rosace <- ImputeData(rosace, key = key, impute.method = "zero")
```

With a complete count matrix in the Assay object, we will normalize the data by either a list of wild-type variants or by the total count at the time point.

```{r}
rosace <- NormalizeData(rosace, key = key,
                        normalization.method = "wt", 
                        wt.var.names = c("_wt"), wt.rm = TRUE)
# rosace <- NormalizeData(rosace, key = key, normalization.method = "total")
```

After the Assay objects are all normalized, we can integrate three replicates into an Assays object stored in the "combined.assay" slot of the Rosace object.

```{r}
rosace <- IntegrateData(object = rosace, key = key)
GetAssaySetName(rosace)
```

#### Naive Method: simple linear regression (optional)

```{r}
rosace <- runSLR(rosace, name = key, type = "AssaySet")
rosace <- runSLR(rosace, name = paste(key, "1", sep = "_"), type = "Assay")
# rosace <- runSLR(rosace, name = paste(key, "2", sep = "_"), type = "Assay")
# rosace <- runSLR(rosace, name = paste(key, "3", sep = "_"), type = "Assay")
```

#### Process variants' meta data (provide your own function)

Provide your own function for parsing "hgvs" into position, mutation, wildtype, and type of mutation.

```{r}
library("dplyr")
colnames(rosace@var.data)
rosace@var.data <- rosace@var.data %>%
  mutate(tmp = substr(variants, 4, nchar(variants) - 1),
         position = as.numeric(gsub("[[:alpha:]]", "", tmp)),
         wildtype = substr(tmp, 1, 1),
         tmp = substr(tmp, 2, nchar(tmp)),
         mutation = gsub("[[:digit:]]", "", tmp)) %>%
  dplyr::select(-tmp)

func_map <- function(wt, mut) {
  if (nchar(wt) == 0) {
    return("NA")
  }
  
  if (wt == mut) {
    return("synonymous")
  } else if (mut == "del") {
    return("deletion")
  } else {
    return("missense")
  }
}

rosace@var.data <- rosace@var.data %>%
  rowwise() %>%
  mutate(type = func_map(wildtype, mutation)) %>%
  ungroup()
head(rosace@var.data)
```

#### Run ROSACE

There are three ways to run ROSACE model:
1) Provide no position or control column. 
The model would treat each variant independently but with shrinkage on the variance component.
2) Provide position column only.
In addition, the model would group variants together with the amino acid position information provided.
3) Provide both position and control column (recommended).
In addition, the model would group all control variants together into one position label.

For the purpose of vignette we shrink down the number of variants to 100 with bruteforce before running ROSACE as running the full dataset might take an hour. 

```{r message=FALSE}
rosace@assays$`1SM73_1`@norm.counts <- rosace@assays$`1SM73_1`@norm.counts[1:100, ]
rosace@assays$`1SM73_1`@norm.var.names <- rosace@assays$`1SM73_1`@norm.var.names[1:100]
rosace <- RunRosace(object = rosace,
                    name = "1SM73_1",
                    type = "Assay",
                    savedir = "../tests/results/stan/assay/", 
                    pos.col = "position", 
                    ctrl.col = "type",
                    ctrl.name = "synonymous")

rosace@assay.sets$`1SM73`@raw.counts <- rosace@assay.sets$`1SM73`@raw.counts[1:100, ]
rosace@assay.sets$`1SM73`@combined.counts <- rosace@assay.sets$`1SM73`@combined.counts[1:100, ]
rosace@assay.sets$`1SM73`@var.names <- rosace@assay.sets$`1SM73`@var.names[1:100]
rosace <- RunRosace(object = rosace,
                    name = "1SM73",
                    type = "AssaySet",
                    savedir = "../tests/results/stan/assayset/",
                    pos.col = "position",
                    ctrl.col = "type", 
                    ctrl.name = "synonymous")

# why is it downloading cmdstan each time? CheckStanSetup()??
```

#### Visualize Results

We provide three options of visualizing the functional score results: heatmap, violin plot, and density plot. A precomputed result on the full dataset is avaliable for demonstration, stored in `../tests/results/stan/assayset_full`. Note: to use the plotting functions, ensure that `scores.data` adheres to the example variants metadata format, with columns including `position`, `wildtype`, `mutation`, `type`, and `score`.
TO DO: should computed scores be stored in the rosace scores slot?
```{r}
# prepare scores dataframe for plotting
score<-read_tsv(file.path("../tests/results/stan/assayset_full/scores_1SM73.tsv"))$mean
scores.data <- cbind(rosace@var.data[2:nrow(rosace@var.data),], score)

score<-read_tsv(file.path("../tests/results/stan/assayset/scores_1SM73.tsv"))$mean
scores.100 <- cbind(rosace@var.data[2:101,], score)
```


```{r}
# plot score heatmap
scoreHeatmap(data = scores.data, savedir = "../tests/results/stan/assayset_full/plot/", 
             name = "Heatmap_1SM73", savepdf = TRUE)

scoreHeatmap(data = scores.100, pos.step = 1, x.text = 10 ,y.text = 10, seq.text = 3, 
             c.scale = list(mid = 0,l1=0.2, l3 = 0.2, p1=0.9, p3 = .4, p4 = 0.9), 
             savedir = "../tests/results/stan/assayset/plot/", 
             name = "Heatmap_1SM73_100", savepdf = FALSE, show = TRUE)
```


```{r}
# plot score violin plot
scoreVlnplot(data = scores.data, savedir = "../tests/results/stan/assayset_full/plot/", 
             name = "ViolinPlot_1SM73", savepdf = TRUE, show = FALSE)

scoreVlnplot(data = scores.100, jitter = TRUE, savedir = "../tests/results/stan/assayset/plot/",
             x.text = 10, y.text = 10, ht = 10, wd = 8, 
             name = "ViolinPlot_1SM73_100", savepdf = FALSE, show = TRUE)
```

```{r}
# plot score density
scoreDensity(scores.data, hist = TRUE, savedir =
               "../tests/results/stan/assayset_full/plot/", name = "DensityPlot_1SM73", show = FALSE)

scoreDensity(scores.100, scale.free = TRUE, space.free = TRUE, savedir =
               "../tests/results/stan/assayset/plot/", name = "DensityPlot_1SM73_100", savepdf = FALSE)
```
