---
title: "Issues with installation"
author: "Riley Xin"
date: "2023-10-06"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## General
This guide navigates through some quirks and considerations during the installation of the *rosace* and *cmdstanr* packages, particularly those related to compiler configurations on different operating systems. While installations on the latest MacOS and Linux with Ubuntu have been smooth, we've noted a few issues for installing on the Linux CentOS distribution related to C++ configurations.


## Installing *cmdstanr*
Before installing *rosace*, you will need to install *cmdstanr* first using the code below.
```{r install cmdstanr, eval=FALSE}
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))

# use cmdstanr to install CmdStan, this requires a working C++ toolchain and compiler
library(cmdstanr)
install_cmdstan(cores = 2)

# check that installation is successful by checking the CmdStan version
cmdstan_version()
```


If you are downloading on a Linux platform and encounter errors, it is likely due to requirements for a C++ complier since *stan* is built in C++. According to [stan-dev](https://github.com/stan-dev/stan/wiki/Coding-Style-and-Idioms#supported-cpp-versions-and-compilers), *stan* is tested on Linux with g++ 5 (a compiler for C++). Ensure that your system is equipped with a gcc version > 5 which includes g++. You can check the current version in the terminal using:
```{bash, eval=FALSE}
gcc --version
```


If gcc is not installed or if your version is below the required, you'll need to install or upgrade it, following guidelines specific to your Linux distribution. You may check the distribution information in the terminal using:
```{bash, eval=FALSE}
lsb_release -a
```


## Installing *configr*
Note: this problem is specific to Linux with CentOS. 
When you install *rosace* with `devtools::install("rosace")`, it will automatically import required packages and attempt to download those that are not installed. You may see an error `C++17 standard requested but CXX17 is not defined` when downloading the package configr. This error arises from the dependency *RcppTOML* from *configr* that also requires a c++ compiler. 


The error has to do with R's configuration for compiling C++17 code. Sometimes the default compiler is too old or it doesn't handle certain situations properly. You can check what your R uses to compile C++ code in the terminal using: 
```{bash, eval=FALSE}
R CMD config CXX
```


To address the above error, you will need to change R's configuration in your R Makevars file in the terminal using a text editor. Below is an example using nano:
```{base, eval = FALSE}
nano ~/.R/Makevars

# once you enter the editing mode in the file, paste the following
CXX17 = g++
CXX17STD = -std=c++17
CXX17FLAGS = -fPIC

# exit and save the file by pressing CTRL + X, followed by Y, and then ENTER.
```


Sometimes the .R directory does not exits, you can create it and the Makevars file with:
```{base, eval = FALSE}
mkdir -p ~/.R
nano ~/.R/Makevars
```


The configuration now tells R to use g++ for compiling C++17 code specifically, using the -std=c++17 standard and -fPIC compiler flags. This should fix the above error and you may initiate *rosace* installation with `devtools::install("rosace")` again.


## Installing *impute*
If you encounter `xx is not available for this version of R`, or `xx is not available for rosace` during the automatic installation, you may try download it manually. For example, the package *impute* is not available on CRAN and can be downloaded using the following: 
```{r, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("impute")
```
