---
title: "Introduction to rosette simulation"
author: "Jingyou Rao"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rosette simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("rosace")
```

## Description


## Example

Again, we are using the oct1 dataset as our example. 

### Read Processed Rosace Object
```{r}
data("oct1_rosace")
key <- "1SM73"
type <- "growth"
```

### Create Naive Score
```{r}
oct1_rosace <- runSLR(oct1_rosace, name = "1SM73_2", type = "Assay")
```

### Create Rosette Object
```{r}
rosette <- CreateRosetteObject(object = oct1_rosace,
                               score.name = "1SM73_2_SLR",
                               pos.col = "position", mut.col = "mutation",
                               ctrl.col = "type", ctrl.name = "synonymous",
                               project.name = "1SM73_2_SLR")
```

### Generate Summary Statistics

#### Dispersion

TODO: description

When calling the function "CreateRosetteObject", the dispersion parameters are automatically inferred. 
```{r}
rosette@disp
rosette@disp.start
```

#### Mutant Group Label

TODO: description

```{r}
hclust <- HclustMutant(rosette, save.plot = FALSE)
rosette <- GenMutLabel(rosette, hclust = hclust, Gmut = 4, save.plot = FALSE)
```

#### Variant Group Label

TODO: description

```{r}
PlotScoreHist(rosette, var.group = FALSE, mut.group = FALSE)
rosette <- GenVarLabel(rosette, Gvar = 2)
PlotScoreHist(rosette, var.group = TRUE, mut.group = TRUE)
```

#### Weight of variant group within mutant group

TODO: description

```{r}
rosette <- PMVCountDist(rosette, pos.missing = 0.2)
```

### Create config for simulation

```{r eval=FALSE}
cfg <- CreateConfig(rosette,
                    n.sim = 2, save.sim = "tests/results/sim", type.sim = "growth",
                    n.rep = 3, n_round = 3,
                    null.var.group = 'var1', wt.effect = -2,
                    seq.shrink = 1.2, seq.depth = 100,
                    lib.shrink = 2,
                    var.shrink = 1, pos.flag = TRUE,
                    mode.sim = "clean")
```

### Run simulation

```{r eval=FALSE}
runRosette(config = cfg, save.tsv = TRUE, save.rosace = TRUE, save.enrich2 = TRUE)
```






